---
title: "data_manipulation_9.18"
author: "Kennedy Wade"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```

lets import the datasets 

```{r}
litters_df = 
    read_csv("./data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
    janitor::clean_names(litters_df)

pups_df = 
    read_csv("./data_import_examples/FAS_pups.csv", na = c("NA", "."), skip = 3)

pups_df = 
    janitor::clean_names(pups_df)

```

## `select`

`select` selects varibales from my dataframe

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range 

```{r}
select(litters_df, group:gd_of_birth)
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting 
```{r}
select(litters_df, group, starts_with("gd"), group)
select(litters_df, group, litter_number, everything())

relocate(litters_df, group, litter_number)
```

rename by selecting 

```{r}
select(litters_df,GROUP = group,everything())
rename(litters_df,GROUP = group)
```

what if i only select one variable...?

```{r}
select(litters_df, group)

pull(litters_df,group)
```

Learning assessment code

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## 'filter`

use filter () to filter out rows 

```{r}
filter(litters_df,gd_of_birth ==20)
```


look for more than 5 pups alive

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive <= 5)
```

look for non - matches

```{r}
filter(litters_df, pups_born_alive != 6)
```

filter out missing rows with `drop_na()`

```{r}
drop_na(litters_df)
drop_na(litters_df , gd0_weight)
```

Learning assessment 2

```{r}
filter(pups_df, sex == 1)
filter(pups_df, sex == 2, pd_walk < 11)
```

##`mutate`

use `mutate()` to create or modify variables.

```{r}
mutate(litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

## `arrange`

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, group, pups_born_alive)
```

## pipes

```{r}
litters_df = 
   read_csv("data_import_examples/FAS_litters.csv", na = c("NA",".","")) |>
janitor::clean_names() |>
select(group, litter_number, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```

learning assessment 

```{r}
pups_df = 
    read_csv("./data_import_examples/FAS_pups.csv", na = c("NA", "."), skip = 3) |>
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(pd_pivot_gt7 = pd_pivot > 7)
```














